Predicting the conformational preferences of proteins using a physics-based free energy method

### 2013-06-17 Paper/paper.tex
correction_June17

The GA and GB sequences are nearly identical, so there are some common features observed for all sequences.
The experimental observations classified the protein into two parts: Amino acids 9--51 are fully structured in both
folds, whereas residues 1--8 and 52--56 are unstructured in the $\alpha$ fold, but form $\beta$-strands in the $\beta$
fold. Most of the amino acid residues in the 
%structured?
%\Arijit {yes}
region 1--8 have negative per-residue free energies, which means that these
residues favor the $\beta$ structure. Roles of some other important residues in stabilizing either the $\alpha$ or
$\beta$ conformation are summarized in Figure S1.

@@

We performed a similar calculation for target T0560 with two models from the group "Splicer''. The remaining three
models were discarded as they were too similar to the rest of the models. Again, we can correctly identify the native
state, and our calculated free energy based ranking matches well with GDT-TS (Figure S2 in the supporting information).

@@

We discussed how the per-residue free energy can reveal mechanistic detail behind the
conformational preference of chameleon sequences. Here, we applied the same method to
see if the per-residue free energy can help us identify problematic residues in structure predictions. We chose
CASP target T0569 and compared the experimental NMR structure with the best predicted model (GDT-TS=78; predicted by the
"Mufold'' group). The confinement method predicts that the experimental structure is more stable by 20 kcal/mol (Figure
S2).
%The annoying rule is that "side chain" is correct for the noun usage, and side-chain for the adjective.
The confinement method clearly identified the two hydrophobic residues V59 and I61 as destabilizing the predicted model
relative to the experimental structure (Figure~\ref{fig:T0569_per_residue} and Figure S3). The side chains of these
hydrophobic residues are oriented towards the protein hydrophobic core in the native NMR structure but are oriented towards
the exterior of the protein, exposing them to solvent, in the model. These residues are part of a beta-sheet in the
experimental structure, but because of their side-chain orientation, the corresponding beta-sheet becomes disordered in
the predicted model (Figure~\ref{fig:T0569_per_residue} and Figure S4). There is also a large difference around K76,
which forms a salt-bridge with D11 in the predicted model, but not in the experimental structure. This suggests that
salt-bridge interactions are too favorable for the combination of forcefield and implicit solvent model we use, which
has been a problem noted in the past \cite{Roe2007}.

@@

Despite success for most of the systems we studied, there were few failures, especially when the GDT-TS scores of the
compared structures were very close. One example is Target T0538, where we compared the experimental structure with three
models (Model 1: "PconsR''---GDT-TS=96; Model 2: "Shell''---GDT-TS=90; Model 3: "FOLDIT''---GDT-TS=86). Contrary to
our expectation, the confinement method predicts that computer generated Model 1 is more stable than the crystal
structure (Figure S5). Per-residue free energy calculations (not shown) show that despite only small variations at the
backbone level, the side chains are oriented in very different ways (Figure~\ref{fig:T0538compare}), giving rise to
large differences in the stabilization of certain residues. In particular, some of the differences arise from different
salt bridge patterns and certain flexible polar residues exposed to the surface. This unexpected result shows that the
confinement method is very sensitive to local interactions (including side chain reorientation) and may indicate issues
with the forcefield and implicit solvent models used in the calculation.

\begin{figure}
\begin{center}
\includegraphics[width=4.9 in,height=4.0 in]{Target_538_compare.pdf}
\end{center}
\caption{Small differences in side-chain orientation can cause large differences in free energy. We compare the
    experimental structure and a computer generated model for CASP target T0538. (A) Two salt bridges between residues
    E35-R32 and E28-K24 exist in the native structure. These are absent in the computer generated model structure.
    Instead, (B) a new salt bridge is formed between residues R32-E28 in model. (C) Another salt bridge between residues
    R26-E50 in the experimental structure. (D) In comparison to the native structure, orientation of F51 differs in the
    model.}
\label{fig:T0538compare}
\end{figure}

>>>>

The GA and GB sequences are nearly identical, so there are some common features observed for all sequences.
The experimental observations classified the protein into two parts: Amino acids 9--51 are fully structured in both
folds, whereas residues 1--8 and 52--56 are unstructured in the $\alpha$ fold, but form $\beta$-strands in the $\beta$
fold. Most of the amino acid residues in the 
%structured?
%\Arijit {yes}
region 1--8 have negative per-residue free energies, which means that these
residues favor the $\beta$ structure. Roles of some other important residues in stabilizing either the $\alpha$ or
$\beta$ conformation are summarized in Figure S1 and S2-S7.

@@

We performed a similar calculation for target T0560 with two models from the group "Splicer''. The remaining three
models were discarded as they were too similar to the rest of the models. Again, we can correctly identify the native
state, and our calculated free energy based ranking matches well with GDT-TS (Figure S8 in the supporting information).

@@

We discussed how the per-residue free energy can reveal mechanistic detail behind the
conformational preference of chameleon sequences. Here, we applied the same method to
see if the per-residue free energy can help us identify problematic residues in structure predictions. We chose
CASP target T0569 and compared the experimental NMR structure with the best predicted model (GDT-TS=78; predicted by the
"Mufold'' group). The confinement method predicts that the experimental structure is more stable by 20 kcal/mol (Figure
S2).
%The annoying rule is that "side chain" is correct for the noun usage, and side-chain for the adjective.
The confinement method clearly identified the two hydrophobic residues V59 and I61 as destabilizing the predicted model
relative to the experimental structure (Figure~\ref{fig:T0569_per_residue} and Figure S9). The side chains of these
hydrophobic residues are oriented towards the protein hydrophobic core in the native NMR structure but are oriented towards
the exterior of the protein, exposing them to solvent, in the model. These residues are part of a beta-sheet in the
experimental structure, but because of their side-chain orientation, the corresponding beta-sheet becomes disordered in
the predicted model (Figure~\ref{fig:T0569_per_residue} and Figure S10). There is also a large difference around K76,
which forms a salt-bridge with D11 in the predicted model, but not in the experimental structure. This suggests that
salt-bridge interactions are too favorable for the combination of forcefield and implicit solvent model we use, which
has been a problem noted in the past \cite{Roe2007}.

@@

Despite success for most of the systems we studied, there were few failures, especially when the GDT-TS scores of the
compared structures were very close. One example is Target T0538, where we compared the experimental structure with three
models (Model 1: "PconsR''---GDT-TS=96; Model 2: "Shell''---GDT-TS=90; Model 3: "FOLDIT''---GDT-TS=86). Contrary to
our expectation, the confinement method predicts that computer generated Model 1 is more stable than the crystal
structure (Figure S11). Per-residue free energy calculations (not shown) show that despite only small variations at the
backbone level, the side chains are oriented in very different ways (Figure~\ref{fig:T0538compare}), giving rise to
large differences in the stabilization of certain residues. In particular, some of the differences arise from different
salt bridge patterns and certain flexible polar residues exposed to the surface. This unexpected result shows that the
confinement method is very sensitive to local interactions (including side chain reorientation) and may indicate issues
with the forcefield and implicit solvent models used in the calculation.

\begin{figure}
\begin{center}
\includegraphics[width=4.8in]{Target_538_compare.pdf}
\end{center}
\caption{Small differences in side-chain orientation can cause large differences in free energy. We compare the
    experimental structure and a computer generated model for CASP target T0538. Two salt bridges between residues
    E35-R32 and E28-K24 exist in the native structure. These are absent in the computer generated model structure.
    Instead, a new salt bridge is formed between residues R32-E28 in model. } 
\label{fig:T0538compare}
\end{figure}
